---
title: "MITII Day 4"
author: "Erick Matsen and Trevor Bedford"
date: "April 24, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R setup

Install [apTreeshape](https://cran.r-project.org/web/packages/apTreeshape/index.html).
```{r eval=FALSE} 
install.packages("apTreeshape")
```

## Doing sequence alignment

First we need to download the raw sequence data.
Visit the [data page for this exercise](https://github.com/blab/mitii/tree/master/phylogenetics/data) and download the data sets.
If you are having trouble with that, simply click [here](https://raw.githubusercontent.com/blab/mitii/master/phylogenetics/data/HIV1_REF_2010_gag_DNA.fasta) and save that page to a file.

Next, visit <http://wasabiapp.org>. 
Click on `Data -> Import` to upload your data.
To align, click on `Tools -> PRANK aligner`.

Once you have your aligned sequences, you can export them using `Data -> Export`. 
*Important:* export your sequences in `phylip` format.


## Building a tree

We will do tree building using <http://www.atgc-montpellier.fr/phyml/>, the ATGC bioinformatics platform.
Upload your data, enter your email address, and hit `Execute`.

Once that completes, hit `Click to visualize your infered tree`.
You can interactively play with how the tree is displayed, then download your tree in [Newick format](https://en.wikipedia.org/wiki/Newick_format).


## Playing with your tree in R

```{r, message=FALSE}
library(apTreeshape)
```

Load the tree into R with the `read.tree` function of [ape](https://cran.r-project.org/web/packages/ape/):
```{r}
t = read.tree('~/file.nwk')
t
```

We can see that the tree is unrooted, which is the norm for doing phylogenetic analysis with reversible models.

The next step is to root the tree at a specific leaf:
```{r}
r = root(t, "Ref.A1.UG.92.92UG037.AB253429", resolve.root = TRUE)
r
```

```{r}
colless(as.treeshape(r))
```