---
title: "B cell analysis with IMGT and bcRep"
author: "Erick Matsen"
date: "May 2, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(bcRep)
```

## Getting data from IMGT

We assume that you have run your data through [IMGT HighV-QUEST](http://imgt.org/HighV-QUEST/login.action).

I've done that for you for samples from the [Rubelt et al](https://www.nature.com/articles/ncomms11112) twins study and put them in a `.zip` archive in the `bcells/data` directory.
Please unzip them in place.

The samples are organized into twin pairs, with 1A/1B being one twin pair, and 2A/2B being another.
All of them are from naive sorts.
Our challenge will be to see if pairs of twins have more similar gene usage than non-twin pairs.

Take a look at the files and get acquainted with their contents.
They're complicated!


## Using bcRep

Parsing these files on our own would be a pain.
Luckily there is an R package for that, called [bcRep](https://cran.r-project.org/web/packages/bcRep).

First we get our data using the `readIMGT` function:

```{r, results="hide"}
aa_1a = readIMGT("~/rubelt/rubelt_naive_1A/5_AA-sequences.txt")
aa_1b = readIMGT("~/rubelt/rubelt_naive_1B/5_AA-sequences.txt")
aa_2a = readIMGT("~/rubelt/rubelt_naive_2A/5_AA-sequences.txt")
aa_2b = readIMGT("~/rubelt/rubelt_naive_2B/5_AA-sequences.txt")
```

Here we can see that the `readIMGT` function has done a nice job of parsing the file.
```{r}
head(aa_1a, n=7)
```
But see that last part, with the `, or`? 
The IMGT output isn't really meant to be parsed by a computer easily, it's meant to be read by humans.
Luckily we can use some tools from bcRep to handle this.

First we need to get all of the `V_GENE_and_allele` rows into a list.
We can do that like so.
```{r}
data = list(aa_1a, aa_1b, aa_2a, aa_2b)
names = list("aa_1a", "aa_1b", "aa_2a", "aa_2b")
v_genes = lapply(data, function(x) x$V_GENE_and_allele)
```

Now we can make the comparison:
```{r}
v_gene_comp = compare.geneUsage(gene.list = v_genes, level = "subgroup", names = names)
v_gene_comp
```
What do you think?

## Exercises

All exercises use the [bcRep documentation](https://cran.r-project.org/web/packages/bcRep/).

1. What is `compare.geneUsage` actually reporting?
2. Find and apply a plotting function to plot the results in `v_gene_comp`.
3. Perform another comparison using this data set.

## Next steps

If you want to learn more about R, here are some [good resources](https://www.rstudio.com/online-learning/#R) to get you started.